---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Loading and preprocessing the data

```{r}
setwd("C:/Users/Harshada Sasturkar/Desktop/RepData_PeerAssessment1")
activity<-read.csv("./activity/activity.csv")
```

The interval column in activity data frame represents time as integer. To distinguish between hours and minutes I created a new column activity$hh.mm where interval is stored in 'hh.mm' format. Also dates are converted into 'Date' format.

```{r}
activity$hh.mm<-activity$interval/100
activity$date<-as.Date(activity$date)
str(activity)
```


## What is mean total number of steps taken per day?

### 1. Total number of steps taken per day:
 Observations in which no measurements of steps were taken are excluded

```{r}
step<-subset(activity,!is.na(activity$steps))
tsteps<-tapply(step$steps,step$date,sum)
tsteps
```

### 2. Histogram of the total number of steps taken each day:

```{r}
 hist(tsteps,xlab="total number of steps",ylab="number of days",main="Histogram of total number of steps taken each day",col="orange")
```

### 3. Mean and median of the total number of steps taken per day:

```{r}
mean(tsteps)
median(tsteps)
```

## What is the average daily activity pattern?
### 1. Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

Average number of steps grouped by time interval, averaged across all days:

```{r}
intsteps<-tapply(activity$steps,activity$hh.mm,mean,na.rm=TRUE)
```

Time series plot:

```{r}
plot(names(intsteps),intsteps,type="l",xlab="time interval (hours)",ylab="number of steps",main="Time series plot")
```

### 2. The 5-minute interval, on average across all the days in the dataset, containing the maximum number of steps:

```{r}
which.max(intsteps)
```

i.e. 8 hr 35 min is that time interval (at index 104)

## Imputing missing values

### 1. Total number of missing values in the dataset:

```{r}
 sum(is.na(activity))
```

### 2. filling in all of the missing values in the dataset:

I've used median values of steps for each time interval.

```{r}
medianint<-tapply(step$steps,step$interval,median)
medianint
```

(step is the data frame without missing value rows) 

### 3. A new dataset that is equal to the original dataset but with the missing data filled in:

```{r}
activity1<-activity
while(sum(is.na(activity1))!=0)
{activity1$steps[which(is.na(activity1))]<-medianint[which(activity1$interval[which(is.na(activity1))]==names(medianint))]}
```

The intervals of missing value rows in new data set (activity1) are matched with the intervals(names) of medianint and the values are filled accordingly.

### 4. Histogram of the total number of steps taken each day:

```{r}
tsteps1<-tapply(activity1$steps,activity1$date,sum)
hist(tsteps1,xlab="total number of steps",ylab="number of days",main="Histogram of total number of steps taken each day",col="pink")
```

### 5. Mean and median total number of steps taken per day:

```{r}
mean(tsteps1)
median(tsteps1)
```

These values differ from the previous ones (mean=10766.19,median=10765). From the histogram it is seen that imputing missing values lead to rise in number of days having total number of steps between 0-5000. Also the mean and median are affected by increased number of observations which weren't previously considered. 

## Are there differences in activity patterns between weekdays and weekends?

### 1. Creating new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day:

```{r}
activity1$wdays <- factor(ifelse(weekdays(activity1$date)=="Saturday"|weekdays(activity1$date)=="Sunday", "weekend", "weekday"))
```

### 2. A panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis):

```{r}
awday<-subset(activity1,activity1$wdays=="weekday")
awend<-subset(activity1,activity1$wdays=="weekend")
intsteps1<-tapply(awday$steps,awday$hh.mm,mean)
intsteps2<-tapply(awend$steps,awend$hh.mm,mean)
par(mfrow=c(1,2))
plot(names(intsteps1),intsteps1,type="l",xlab="time interval (hours)",ylab= "number of steps", main="Weekday")
plot(names(intsteps2),intsteps2,type="l",xlab="time interval (hours)",ylab= "number of steps", main="Weekend")
```
